---
title: "Statistical Analysis of Fire Incidents"
author: "Danielle Mirzaeian, Nahom Fisseha"
date: '2022-06-22'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Many calls are made to fire departments all the time. Whether it is a false alarm or not, it is important for first responders to take care of fire incidents as soon as possible.

There are all kinds of fire incidents, some not even about fires. For example someone could be having a stroke and the local fire department was called to take action on the situation. If the incident is a fire, it could be at all kinds of different properties. Some different properties are small houses, apartment buildings, and highways. The fire could vary in size, like the wildfires in California.

No matter the type of fire incident. It is vital that first responders attend to the situation quickly. Now, there can be some factors that can affect the response time of the first responders.
Some of those factors are the location, the type of property, and the type of the incident.

For more than ten years, the city of Greensboro, NC has been collecting data on fire incidents in their area. Some of this data includes the date of the incident, the response time, the location, and the type of property.

For this project we will be visualizing the fire incidents in Greensboro, NC. Our research questions are 1) What factors affect the response time to fire incidents? 2) How to improve the response time? 3) How does response time by fire emergency services affect the rate of total loss and civilian injuries? 4)Can we build models to forecast the number of fire incidents in certain areas of Greensboro based on the data from 2010 to 2022?

### Loading necessary packages

```{r}
library(readr)
library(lubridate)
library(ggplot2)
library(dslabs)
library(dplyr)
library(ggrepel)
library(ggthemes)
```

### Loading the dataset
```{r}
gso.fire <- read_csv("Greensboro_Fire_Incidents.csv")
```


### Filtering for incidents prior to June 1st 2022
```{r}
gso.fire = 
  gso.fire %>%
  mutate(AlarmDate2 = make_date(Year, Month, Day)) %>%
  filter(AlarmDate2 < '2022-06-01')
```

### Filter for property type 1-2 family dwellings
```{r}
gso.fire = gso.fire %>%
  filter(PropertyUse == '419 - 1 or 2 family dwelling')
```

### Convert to call process time to seconds and add to gso.fire
```{r}
gso.fire = gso.fire %>%
  mutate(call_process_period = lubridate::hms(CallProcessingTime), 
         call_process_seconds = period_to_seconds(call_process_period))
```

### Convert response time to seconds and add to gso.fire
```{r}
gso.fire = gso.fire %>%
  mutate(response_time_period = lubridate::hms(ResponseTime),
         response_time_seconds = period_to_seconds(response_time_period))
```

### Convert total response time to seconds and add to gso.fire
```{r}
gso.fire = gso.fire %>%
  mutate(total_response_period = lubridate::hms(TotalResponseTime),
         total_response_seconds = period_to_seconds(total_response_period))
```

### Cutoff point for total response time
```{r}
cutoff.total_response_time = gso.fire %>% select(total_response_seconds) %>% 
  filter(!is.na(total_response_seconds)) %>%
  summarize(c = quantile(total_response_seconds, .75) + 3 * IQR(total_response_seconds)) %>%
  pull(c)
```

## Histogram of distribution of total response time
```{r}
gso.fire %>%
  filter(total_response_seconds < cutoff.total_response_time) %>%
  ggplot(aes(x = total_response_seconds)) +
  geom_histogram(aes(x = total_response_seconds, fill = "red"), show.legend = FALSE, color = "black") +
  xlab("Total Response Time (Seconds)") +
  ylab("Count") +
  ggtitle("Distribution of Total Response Time") +
  theme_economist()
```

## Histogram of frequency of fire alarms by alarm hour
```{r}
gso.fire %>%
  ggplot(aes(x = AlarmHour)) +
  geom_bar(color = "black", fill = "pink") +
  scale_x_continuous(breaks = seq(0,23,1)) +
  theme(axis.text.x = element_text(angle = 45)) +
  xlab("Alarm Hour") +
  ylab("Count") +
  ggtitle("Frequency of Fire Alarms by Alarm Hour")
```

## Bar graph of frequency of calls by fire station
```{r}
gso.fire %>%
  count(station) %>%
  mutate(station = reorder(station, n)) %>%
  ggplot(aes(x = station, y = n)) +
  geom_bar(stat = "identity", color = "black", fill = "pink") +
  theme(axis.text.y = element_text(size = 6)) +
  xlab("Fire Station (Station Number)") +
  ylab("Count") +
  ggtitle("Frequency of Calls by Fire Station") +
  coord_flip()
```

## Bar graph of frequency of calls by fire district
```{r}
gso.fire %>%
  count(FireDistrict) %>%
  mutate(FireDistrict = reorder(FireDistrict, n)) %>%
  ggplot(aes(x = FireDistrict, y = n)) +
  geom_bar(stat = "identity", color = "black", fill = "pink") +
  theme(axis.text.y = element_text(size = 6)) +
  xlab("Fire Station (Station Number)") +
  ylab("Count") +
  ggtitle("Frequency of Calls by Fire Station") +
  coord_flip()
```

## Bar graph of frequency of calls by year
```{r}
gso.fire %>%
  count(Year) %>%
  mutate(Year = reorder(Year, n)) %>%
  ggplot(aes(x = Year, y = n)) +
  geom_bar(stat = "identity", color = "black", fill = "light blue") +
  xlab("Year") +
  ylab("Count") +
  ggtitle("Frequency of Calls by Year")
```

## Bar graph of frequency of calls by shift
```{r}
gso.fire %>%
  ggplot(aes(x = shift)) +
  geom_bar(aes(fill = shift), color = "black", show.legend = FALSE) +
  xlab("Shift") +
  ylab("Count") +
  ggtitle("Frequency of Calls by Shift")
```

## Box plot of days of the week vs total response time
```{r}
gso.fire %>%
  filter(total_response_seconds < cutoff.total_response_time) %>%
  ggplot(aes(x = DayOfWeek, y = total_response_seconds, fill = DayOfWeek)) +
  geom_boxplot() +
  xlab("Days of the Week") +
  ylab("Total Response Time (Seconds)") +
  ggtitle("Days of the Week vs Total Repsonse Time") +
  theme_economist()
```

## Box plot of week vs total response time
```{r}
gso.fire %>%
  filter(total_response_seconds < cutoff.total_response_time) %>%
  ggplot(aes(x = Week, y = total_response_seconds, group = Week)) +
  geom_boxplot() +
  xlab("Week") +
  ylab("Total Response Time (Seconds)") +
  ggtitle("Week vs Total Response Time") +
  theme_economist()
```

## Box plot of year vs total response time
```{r}
gso.fire %>%
  filter(total_response_seconds < cutoff.total_response_time) %>%
  ggplot(aes(x = Year, y = total_response_seconds, group = Year)) +
  geom_boxplot(aes(fill = "red"), show.legend = FALSE) +
  xlab("Year") +
  ylab("Total Response Time (Seconds)") +
  ggtitle("Year vs Total Response Time")
```

## Box plot of fire district vs total response time
```{r}
gso.fire %>%
  filter(total_response_seconds < cutoff.total_response_time) %>%
  ggplot(aes(x = FireDistrict, y = total_response_seconds, fill = FireDistrict)) +
  geom_boxplot(show.legend = FALSE) +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Fire District") +
  ylab("Total Response Time (Seconds)") +
  ggtitle("Fire District vs Total Response Time")
```

## Box plot of work shift vs total response time
```{r}
gso.fire %>%
  filter(total_response_seconds < cutoff.total_response_time) %>%
  ggplot(aes(x = shift, y = total_response_seconds, fill = shift)) +
  geom_boxplot() +
  xlab("Work shift") +
  ylab("Total Response Time (Seconds)") +
  ggtitle("Work Shift vs Total Response Time") +
  theme_economist()
```

## Box plot of month vs total response time
```{r}
gso.fire %>%
  filter(total_response_seconds < cutoff.total_response_time) %>%
  ggplot(aes(x = Month, y = total_response_seconds, fill = Month)) +
  geom_boxplot(show.legend = FALSE) +
  xlab("Month") +
  ylab("Total Response Time (Seconds)") +
  ggtitle("Month vs Total Response Time")
```

## Box plot of station vs total response time
```{r}
gso.fire %>%
  filter(total_response_seconds < cutoff.total_response_time) %>%
  ggplot(aes(x = station, y = total_response_seconds)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Station") +
  ylab("Total Response Time (Seconds)") +
  ggtitle("Station vs Total Response Time")
```

## Box plot of alarm hour vs total response time
```{r}
gso.fire %>%
  filter(total_response_seconds < cutoff.total_response_time) %>%
  ggplot(aes(x = AlarmHour, y = total_response_seconds, group = AlarmHour)) +
  geom_boxplot() +
  xlab("Alarm Hour") +
  ylab("Total Response Time (Seconds)") +
  ggtitle("Alarm Hour vs Total Response Time")
```

## Box plot of number of alarms vs total response time
```{r}
gso.fire %>%
  filter(total_response_seconds < cutoff.total_response_time) %>%
  ggplot(aes(x = NumberOfAlarms, y = total_response_seconds, group = NumberOfAlarms)) +
  geom_boxplot() +
  xlab("Number of Alarms") +
  ylab("Total Response Time (Seconds)") +
  ggtitle("Number of Alarms vs Total Response Time")
```

## Box plot of nature code vs total response time
```{r}
gso.fire %>%
  filter(total_response_seconds < cutoff.total_response_time) %>%
  ggplot(aes(x =  NatureCode, y = total_response_seconds)) +
  theme(axis.text.x = element_text(angle = 90)) +
  geom_boxplot() +
  xlab("Nature Code") +
  ylab("Total Response Time (Seconds)") +
  ggtitle("Nature Code vs Total Response Time")
```

## Scatter plot of Total Staff on Incident vs Total Response
```{r}
gso.fire %>%
  filter(total_response_seconds < cutoff.total_response_time) %>%
  ggplot(aes(x = total_response_seconds, y = TotalStaffOnIncident, fill = total_response_seconds)) +
  geom_point(stat = "identity", show.legend = FALSE) +
  ylab("Total Staff On Incident") +
  xlab("Total Response Time") +
  ggtitle("Total Staff on Incident vs Total Response Time") +
  theme_economist()
```

## Scatter plot of Civilian Injuries vs Total Response Time
```{r}
gso.fire %>%
  filter(total_response_seconds < cutoff.total_response_time) %>%
  ggplot(aes(x = total_response_seconds, y = CivilianInjuries, fill = total_response_seconds)) +
  geom_point(stat = "identity", show.legend = FALSE) +
  ylab("Civilian Injuries") +
  xlab("Total Response Time") +
  ggtitle("Civilian Injuries vs Total Response Time") +
  theme_economist() 
```

## Scatter Plot for Civilian Fatalities vs Total Response Time
```{r}
gso.fire %>%
  filter(total_response_seconds < cutoff.total_response_time) %>%
  ggplot(aes(x = total_response_seconds, y = CivilianFatalities, fill = total_response_seconds)) +
  geom_point(stat = "identity", show.legend = FALSE) +
  ylab("Civilian Fatalities") +
  xlab("Total Response Time") +
  ggtitle("Civilian Fatalities vs Total Response Time") + 
  theme_economist()
```

## Scatter Plot for Fire Service Injuries vs Total Response Time
```{r}
gso.fire %>%
  filter(total_response_seconds < cutoff.total_response_time) %>%
  ggplot(aes(x = total_response_seconds, y = FireServiceInjuries, fill = total_response_seconds)) +
  geom_point(stat = "identity", show.legend = FALSE) +
  ylab("Fire Service Injuries") +
  xlab("Total Response Time") +
  ggtitle("Fire Service Injuries Injuries vs Total Response Time") +
  theme_economist()
```

## Scatter Plot for Fire Service Fatalities vs Total Response Time
```{r}
gso.fire %>%
  filter(total_response_seconds < cutoff.total_response_time) %>%
  ggplot(aes(x = total_response_seconds, y = FireServiceFatalities, fill = TotalResponseTime)) +
  geom_point(stat = "identity", show.legend = FALSE) +
  ylab("Fire Service Fatalities") +
  xlab("Total Response Time") +
  ggtitle("Fire Service Fatalities vs Total Response Time") +
  theme_economist() 
```



