---
title: "Statistical Analysis of Fire Incidents"
author: "Danielle Mirzaeian, Nahom Fisseha"
date: '2022-06-22'
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Many calls are made to fire departments all the time. Whether it is a false alarm or not, it is important for first responders to take care of fire incidents as soon as possible.

There are all kinds of fire incidents, some not even about fires. For example someone could be having a stroke and the local fire department was called to take action on the situation. If the incident is a fire, it could be at all kinds of different properties. Some different properties are small houses, apartment buildings, and highways. The fire could vary in size, like the wildfires in California.

No matter the type of fire incident. It is vital that first responders attend to the situation quickly. Now, there can be some factors that can affect the response time of the first responders.
Some of those factors are the location, the type of property, and the type of the incident.

For more than ten years, the city of Greensboro, NC has been collecting data on fire incidents in their area. Some of this data includes the date of the incident, the response time, the location, and the type of property. Yet, take note when looking at the data that it goes from July 1, 2010, to Jun 1, 2022. So for the years 2010 and 2022, only data for about half of those years was collected.

For this project we will be visualizing the fire incidents in Greensboro, NC. Our research questions are 1) What factors affect the response time to fire incidents? 2) How to improve the response time? 3) How does response time by fire emergency services affect the rate of total loss and civilian injuries? 4)Can we build models to forecast the number of fire incidents in certain areas of Greensboro based on the data from 2010 to 2022? 5) what Dynamic models can we make for forecasting fire incidents?

## Related Literature

file:///Users/nah8m/Downloads/KCK.andCorcoranJ.2017.Modellingresidentialfireincidentresponsetimes.pdf

file:///Users/nah8m/Downloads/Relationshipbetweenfireserviceresponseandfiregrowth.pdf

## Method
### Exploratory Data Analysis
The first step in our research project was exploratory data analysis. We investigated the data looking for patterns and outliers. This was carried out by making various graphical representations of the data.

#### Loading necessary packages

```{r, message=FALSE, warning=FALSE}
library(readr)
library(lubridate)
library(ggplot2)
library(dslabs)
library(dplyr)
library(ggrepel)
library(ggthemes)
```

#### Data Preprocessing
```{r, message=FALSE, warning=FALSE}
gso.fire <- read_csv("Greensboro_Fire_Incidents.csv")
```


#### Filtering for incidents prior to June 1st 2022
```{r}
gso.fire = 
  gso.fire %>%
  mutate(AlarmDate2 = make_date(Year, Month, Day)) %>%
  filter(AlarmDate2 < '2022-06-01')
```

#### Filter for property type 1-2 family dwellings
```{r}
gso.fire = gso.fire %>%
  filter(PropertyUse == '419 - 1 or 2 family dwelling')
```

#### Convert to call process time to seconds and add to gso.fire
```{r}
gso.fire = gso.fire %>%
  mutate(call_process_period = lubridate::hms(CallProcessingTime), 
         call_process_seconds = period_to_seconds(call_process_period))
```

#### Convert response time to seconds and add to gso.fire
```{r, message=FALSE, warning=FALSE}
gso.fire = gso.fire %>%
  mutate(response_time_period = lubridate::hms(ResponseTime),
         response_time_seconds = period_to_seconds(response_time_period))
```

#### Convert total response time to seconds and add to gso.fire
```{r, message=FALSE, warning=FALSE}
gso.fire = gso.fire %>%
  mutate(total_response_period = lubridate::hms(TotalResponseTime),
         total_response_seconds = period_to_seconds(total_response_period))
```

#### Cutoff point for response time
```{r}
cutoff.response_time = gso.fire %>% select(response_time_seconds) %>%
  filter(!is.na(response_time_seconds)) %>%
  summarize(c = quantile(response_time_seconds, .75) + 3 * IQR(response_time_seconds)) %>%
  pull(c)
```

## Graphs

### Histogram of distribution of response time
```{r, message=FALSE, warning=FALSE}
gso.fire %>%
  filter(response_time_seconds < cutoff.response_time) %>%
  ggplot(aes(x = response_time_seconds, fill = "red")) +
  geom_histogram(show.legend = FALSE, color = "black") +
  xlab("Response Time (Seconds)") +
  ylab("Count") +
  ggtitle("Distribution of Response Time") +
  theme_economist()
```

For the histogram above, we learned that the typical response time falls between about 100 seconds to 400 seconds and that the distribution is normal.

### Bar graph of frequency of fire alarms by alarm hour
```{r}
gso.fire %>%
  ggplot(aes(x = AlarmHour)) +
  geom_bar(color = "black", fill = "pink") +
  scale_x_continuous(breaks = seq(0,23,1)) +
  theme(axis.text.x = element_text(angle = 45)) +
  xlab("Alarm Hour") +
  ylab("Count") +
  ggtitle("Frequency of Fire Alarms by Alarm Hour")
```

From this bar graph, it can be seen that the frequency of calls to the fire stations are less during the middle of the night and increase as people start waking up.

### Bar graph of frequency of calls by fire station
```{r}
gso.fire %>%
  count(station) %>%
  mutate(station = reorder(station, n)) %>%
  ggplot(aes(x = station, y = n)) +
  geom_bar(stat = "identity", color = "black", fill = "pink") +
  theme(axis.text.y = element_text(size = 6)) +
  xlab("Fire Station (Station Number)") +
  ylab("Count") +
  ggtitle("Frequency of Calls by Fire Station") +
  coord_flip()
```

### Bar graph of frequency of calls by fire district
```{r}
gso.fire %>%
  count(FireDistrict) %>%
  mutate(FireDistrict = reorder(FireDistrict, n)) %>%
  ggplot(aes(x = FireDistrict, y = n)) +
  geom_bar(stat = "identity", color = "black", fill = "pink") +
  theme(axis.text.y = element_text(size = 6)) +
  xlab("Fire Station (Station Number)") +
  ylab("Count") +
  ggtitle("Frequency of Calls by Fire Station") +
  coord_flip()
```

### Bar graph of frequency of calls by year
```{r}
gso.fire %>%
  count(Year) %>%
  ggplot(aes(x = Year, y = n)) +
  geom_bar(stat = "identity", fill = "pink", color = "black") +
  scale_x_continuous(breaks = seq(2010,2022,1)) +
  xlab("Year") +
  ylab("Count") +
  ggtitle("Frequency of Calls by Year")
```


### Bar graph of frequency of calls by shift
```{r}
gso.fire %>%
  ggplot(aes(x = shift)) +
  geom_bar(show.legend = FALSE, fill = "pink", color = "black") +
  xlab("Shift") +
  ylab("Count") +
  ggtitle("Frequency of Calls by Shift")
```

### Bar graph of frequency of calls by days of the week
```{r}
gso.fire %>%
  ggplot(aes(x = DayOfWeek)) +
  geom_bar(fill = "pink", color = "black") +
  xlab("Days of the Week") +
  ylab("Count") +
  ggtitle("Frequency of Calls by Days of the Week")
```

### Bar graph of frequency of calls by week
```{r}
gso.fire %>%
  ggplot(aes(x = Week)) +
  geom_bar(fill = "pink", color = "black") +
  scale_x_continuous(breaks = seq(1,53,1)) +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Week") +
  ylab("Count") +
  ggtitle("Frequency of Calls by Week")
```

### Bar graph of frequency of calls by month 
```{r}
gso.fire %>%
  ggplot(aes(x = Month)) +
  geom_bar(fill = "pink", color = "black") +
  xlab("Month") +
  ylab("Count") +
  ggtitle("Frequency of Calls by Month")
```


### Box plot of week vs response time
```{r}
gso.fire %>%
  filter(response_time_seconds < cutoff.response_time) %>%
  ggplot(aes(x = Week, y = response_time_seconds, group = Week)) +
  geom_boxplot() +
  scale_x_continuous(breaks = seq(1,53,1)) +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Week") +
  ylab("Response Time (Seconds)") +
  ggtitle("Week vs Response Time")
```


### Box plot of station vs total response time
```{r}
gso.fire %>%
  filter(response_time_seconds < cutoff.response_time) %>%
  ggplot(aes(x = station, y = response_time_seconds)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Station") +
  ylab("Response Time (Seconds)") +
  ggtitle("Station vs Response Time")
```


### Box plot of nature code vs total response time
```{r}
gso.fire %>%
  filter(response_time_seconds < cutoff.response_time) %>%
  ggplot(aes(x = NatureCode, y = response_time_seconds)) +
  theme(axis.text.x = element_text(angle = 90)) +
  geom_boxplot() +
  xlab("Nature Code") +
  ylab("Response Time (Seconds)") +
  ggtitle("Nature Code vs Response Time")
```


### Scatter plot of Total Staff on Incident vs Response Time
```{r}
gso.fire %>%
  filter(response_time_seconds < cutoff.response_time) %>%
  ggplot(aes(x = response_time_seconds, y = TotalStaffOnIncident)) +
  geom_point(stat = "identity", show.legend = FALSE) +
  ylab("Total Staff On Incident") +
  xlab("Response Time (Seconds)") +
  ggtitle("Total Staff on Incident vs Response Time") +
  theme_economist()
```


### Scatter plot of Civilian Injuries vs Response Time
```{r}
gso.fire %>%
  filter(response_time_seconds < cutoff.response_time) %>%
  ggplot(aes(x = response_time_seconds, y = CivilianInjuries)) +
  geom_point(stat = "identity", show.legend = FALSE) +
  ylab("Civilian Injuries") +
  xlab("Response Time (Seconds)") +
  ggtitle("Civilian Injuries vs Response Time") +
  theme_economist() 
```


### Scatter Plot for Civilian Fatalities vs Response Time
```{r}
gso.fire %>%
  filter(response_time_seconds < cutoff.response_time) %>%
  ggplot(aes(x = response_time_seconds, y = CivilianFatalities)) +
  geom_point(stat = "identity", show.legend = FALSE) +
  ylab("Civilian Fatalities") +
  xlab("Response Time (Seconds)") +
  ggtitle("Civilian Fatalities vs Response Time") + 
  theme_economist()
```


### Scatter Plot for Fire Service Injuries vs Response Time
```{r}
gso.fire %>%
  filter(response_time_seconds < cutoff.response_time) %>%
  ggplot(aes(x = response_time_seconds, y = FireServiceInjuries)) +
  geom_point(stat = "identity", show.legend = FALSE) +
  ylab("Fire Service Injuries") +
  xlab("Response Time (Seconds)") +
  ggtitle("Fire Service Injuries Injuries vs Response Time") +
  theme_economist()
```


### Scatter Plot for Fire Service Fatalities vs Response Time
```{r}
gso.fire %>%
  filter(response_time_seconds < cutoff.response_time) %>%
  ggplot(aes(x = response_time_seconds, y = FireServiceFatalities)) +
  geom_point(stat = "identity", show.legend = FALSE) +
  ylab("Fire Service Fatalities") +
  xlab("Response Time (Seconds)") +
  ggtitle("Fire Service Fatalities vs Response Time") +
  theme_economist() 
```


### Box and Violin plot for days of the week vs response time
```{r}
gso.fire %>%
  filter(!is.na(response_time_seconds)) %>%
  filter(response_time_seconds < cutoff.response_time) %>%
  ggplot(aes(x = DayOfWeek, y = response_time_seconds)) +
  geom_violin(trim = FALSE, fill = "light blue") +
  geom_boxplot(width = 0.3, fill = "pink") +
  xlab("Days of the Week") +
  ylab("Response Time (Seconds)") +
  ggtitle("Violin & Box Plot of Response Time Densities by Days of the Week") +
  theme(axis.text.x = element_text(angle = 90), legend.position = "hide")
```


### Box and violin plot for year vs response time
```{r}
gso.fire %>%
  filter(response_time_seconds < cutoff.response_time) %>%
  ggplot(aes(x = Year, y = response_time_seconds, group = Year)) +
  geom_violin(trim = FALSE, fill = "light blue") +
  geom_boxplot(show.legend = FALSE, width = .3, fill = "pink") +
  scale_x_continuous(breaks = seq(2010,2022,1)) +
  xlab("Year") +
  ylab("Response Time (Seconds)") +
  ggtitle("Violin & Box Plots of Response Time Densities by Year")
```


### Box and violin plot for work shift vs response time
```{r}
gso.fire %>%
  filter(response_time_seconds < cutoff.response_time) %>%
  ggplot(aes(x = shift, y = response_time_seconds)) +
  geom_violin(show.legend = FALSE, fill = "light blue") +
  geom_boxplot(width = .5, show.legend = FALSE, fill = "pink") +
  xlab("Work shift") +
  ylab("Response Time (Seconds)") +
  ggtitle("Violin and Box Plot of Response Time Densities by Work Shift")
```


### Box and violin plot for fire district vs response time
```{r, message=FALSE, warning=FALSE}
gso.fire %>%
  filter(response_time_seconds < cutoff.response_time) %>%
  ggplot(aes(x = FireDistrict, y = response_time_seconds)) +
  geom_violin(fill = "light blue") +
  geom_boxplot(show.legend = FALSE, width = .2, fill = "pink") +
  theme(axis.text.x = element_text(angle = 90), legend.position = "hide") +
  xlab("Fire District") +
  ylab("Response Time (Seconds)") +
  ggtitle("Violin and Box Plot of Response Time Densities by Fire District")
```


### Box and violin plot for month vs response time
```{r}
gso.fire %>%
  filter(response_time_seconds < cutoff.response_time) %>%
  ggplot(aes(x = Month, y = response_time_seconds)) +
  geom_violin(show.legend = FALSE, fill = "light blue") +
  geom_boxplot(show.legend = FALSE, width = .3, fill = "pink") +
  xlab("Month") +
  ylab("Response Time (Seconds)") +
  ggtitle("Violin and Box Plot of Response Time Densities by Month")
```


### Box and violin plot for alarm hour vs total response time
```{r}
gso.fire %>%
  filter(response_time_seconds < cutoff.response_time) %>%
  ggplot(aes(x = AlarmHour, y = response_time_seconds, group = AlarmHour)) +
  geom_violin(fill = "light blue") +
  geom_boxplot(width = .3, fill = "pink") +
  scale_x_continuous(breaks = seq(0,23,1)) +
  xlab("Alarm Hour") +
  ylab("Response Time (Seconds)") +
  ggtitle("Violin and Box Plot of Response Time Densities by Alarm Hour")
```


### Box and violin plot of number of alarms vs response time
```{r, message=FALSE, warning=FALSE}
gso.fire %>%
  filter(response_time_seconds < cutoff.response_time) %>%
  ggplot(aes(x = NumberOfAlarms, y = response_time_seconds, group = NumberOfAlarms)) +
  geom_violin(fill = "light blue") +
  geom_boxplot(width = .3, fill = "pink") +
  xlab("Number of Alarms") +
  ylab("Response Time (Seconds)") +
  ggtitle("Violin and Box Plot of Response Time Densities by Number of Alarms")
```

## Time Series Forecasting

#### Filtering data
```{r}
gso.fire.filtered = gso.fire %>%
  filter(response_time_seconds < cutoff.response_time)
```

#### Daily number of fire incidents
```{r}
gso.fire.ts = gso.fire %>%
  count(AlarmDate2)
```


### Modeling response time
```{r}
lm.res.time = lm(response_time_seconds ~ TotalStaffOnIncident + FireDistrict + DayOfWeek + shift + Month, 
                 data = gso.fire.filtered)
```


### Time series of daily number of fire incidents for 2010
```{r}
gso.fire.ts %>%
  filter(AlarmDate2 >= "2010-01-01" & AlarmDate2 <= "2010-12-31") %>%
  ggplot(aes(x = AlarmDate2, y = n)) +
  geom_line() +
  xlab("Alarm Date") +
  ylab("Number of Fire Incidents") +
  ggtitle("Time series of daily number of fire incidents for 2010")
```


### Time series of daily number of fire incidents for 2011
```{r}
gso.fire.ts %>%
  filter(AlarmDate2 >= "2011-01-01" & AlarmDate2 <= "2011-12-31") %>%
  ggplot(aes(x = AlarmDate2, y = n)) +
  geom_line() +
  xlab("Alarm Date") +
  ylab("Number of Fire Incidents") +
  ggtitle("Time series of daily number of fire incidents for 2011")
```


### Time series of daily number of fire incidents for 2012
```{r}
gso.fire.ts %>%
  filter(AlarmDate2 >= "2012-01-01" & AlarmDate2 <= "2012-12-31") %>%
  ggplot(aes(x = AlarmDate2, y = n)) +
  geom_line() +
  xlab("Alarm Date") +
  ylab("Number of Fire Incidents") +
  ggtitle("Time series of daily number of fire incidents for 2012")
```


### Time series of daily number of fire incidents for 2013
```{r}
gso.fire.ts %>%
  filter(AlarmDate2 >= "2013-01-01" & AlarmDate2 <= "2013-12-31") %>%
  ggplot(aes(x = AlarmDate2, y = n)) +
  geom_line() +
  xlab("Alarm Date") +
  ylab("Number of Fire Incidents") +
  ggtitle("Time series of daily number of fire incidents for 2013")
```


### Time series of daily number of fire incidents for 2014
```{r}
gso.fire.ts %>%
  filter(AlarmDate2 >= "2014-01-01" & AlarmDate2 <= "2014-12-31") %>%
  ggplot(aes(x = AlarmDate2, y = n)) +
  geom_line() +
  xlab("Alarm Date") +
  ylab("Number of Fire Incidents") +
  ggtitle("Time series of daily number of fire incidents for 2014")
```


### Time series of daily number of fire incidents for 2015
```{r}
gso.fire.ts %>%
  filter(AlarmDate2 >= "2015-01-01" & AlarmDate2 <= "2015-12-31") %>%
  ggplot(aes(x = AlarmDate2, y = n)) +
  geom_line() +
  xlab("Alarm Date") +
  ylab("Number of Fire Incidents") +
  ggtitle("Time series of daily number of fire incidents for 2015")
```


### Time series of daily number of fire incidents for 2016
```{r}
gso.fire.ts %>%
  filter(AlarmDate2 >= "2016-01-01" & AlarmDate2 <= "2016-12-31") %>%
  ggplot(aes(x = AlarmDate2, y = n)) +
  geom_line() +
  xlab("Alarm Date") +
  ylab("Number of Fire Incidents") +
  ggtitle("Time series of daily number of fire incidents for 2016")
```


### Time series of daily number of fire incidents for 2017
```{r}
gso.fire.ts %>%
  filter(AlarmDate2 >= "2017-01-01" & AlarmDate2 <= "2017-12-31") %>%
  ggplot(aes(x = AlarmDate2, y = n)) +
  geom_line() +
  xlab("Alarm Date") +
  ylab("Number of Fire Incidents") +
  ggtitle("Time series of daily number of fire incidents for 2017")
```


### Time series of daily number of fire incidents for 2018
```{r}
gso.fire.ts %>%
  filter(AlarmDate2 >= "2018-01-01" & AlarmDate2 <= "2018-12-31") %>%
  ggplot(aes(x = AlarmDate2, y = n)) +
  geom_line() +
  xlab("Alarm Date") +
  ylab("Number of Fire Incidents") +
  ggtitle("Time series of daily number of fire incidents for 2018")
```


### Time series of daily number of fire incidents for 2019
```{r}
gso.fire.ts %>%
  filter(AlarmDate2 >= "2019-01-01" & AlarmDate2 <= "2019-12-31") %>%
  ggplot(aes(x = AlarmDate2, y = n)) +
  geom_line() +
  xlab("Alarm Date") +
  ylab("Number of Fire Incidents") +
  ggtitle("Time series of daily number of fire incidents for 2019")
```


### Time series of daily number of fire incidents for 2020
```{r}
gso.fire.ts %>%
  filter(AlarmDate2 >= "2020-01-01" & AlarmDate2 <= "2020-12-31") %>%
  ggplot(aes(x = AlarmDate2, y = n)) +
  geom_line() +
  xlab("Alarm Date") +
  ylab("Number of Fire Incidents") +
  ggtitle("Time series of daily number of fire incidents for 2020")
```


### Time series of daily number of fire incidents for 2021
```{r}
gso.fire.ts %>%
  filter(AlarmDate2 >= "2021-01-01" & AlarmDate2 <= "2021-12-31") %>%
  ggplot(aes(x = AlarmDate2, y = n)) +
  geom_line() +
  xlab("Alarm Date") +
  ylab("Number of Fire Incidents") +
  ggtitle("Time series of daily number of fire incidents for 2021")
```


### Time series of daily number of fire incidents for 2022
```{r}
gso.fire.ts %>%
  filter(AlarmDate2 >= "2022-01-01" & AlarmDate2 <= "2022-12-31") %>%
  ggplot(aes(x = AlarmDate2, y = n)) +
  geom_line() +
  xlab("Alarm Date") +
  ylab("Number of Fire Incidents") +
  ggtitle("Time series of daily number of fire incidents for 2022")
```


#### Monthly number of fire incidents
```{r}
gso.fire.ts.month = gso.fire %>%
  group_by(Year, Month) %>%
  count()
```

